---
title: "Identifying rewired proteins in comparative CF-MS data"
author: "--"
output:
  rmarkdown::html_document:
    theme: yeti
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "~/git/PrInCE-vignettes")
knitr::opts_knit$set(root.dir = "~/git/PrInCE-vignettes")
options(stringsAsFactors = F)
suppressWarnings(suppressPackageStartupMessages({
  library(tidyverse)
  library(magrittr)
  library(PrInCE)
  library(fgsea)
}))
```

```{r echo = FALSE, include=FALSE, message=FALSE, warning=FALSE}
source("R/theme.R")
```

This vignette demonstrates the application of the PrInCE function `calculate_autocorrelation` to identify rewired proteins in CF-MS data acquired from human cell lines before and after treatment with RNAse. 
This data is obtained from Mallam et al., _Cell Reports_ 2019, doi: 10.1016/j.celrep.2019.09.060.
To validate the biological relevance of these rewired proteins, a gene-set enrichment analysis is performed for known RNA-binding proteins from two databases: AtTRACT (Giudice et al., _Database_ 2016; doi: 10.1093/database/baw035) and RBPDB (Cook et al., _Nucleic Acids Research_ 2011; doi: 10.1093/nar/gkq1069). 

First, we will load the data from Mallam et al.
Data from human cell lines was obtained from the PRIDE repository (accession PXD014820 and PXD015406) and re-analyzed using MaxQuant (version 1.6.5.0, with default settings, using MaxLFQ for label-free protein quantitation). 
The MaxQuant output (`proteinGroups.txt` files) was then pre-processed using the script in `R/preprocessing/preprocess-Mallam2019.R`. 

```{r}
chroms = readRDS("data/Mallam2019/chromatograms.rds")
names(chroms)
```

Now, it is straightforward to calculate the autocorrelation for each protein between each pair of conditions.
This function performs the following operations.
First, for each individual protein in turn, the Pearson correlation to all other proteins in the dataset is calculated in each condition separately, yielding two vectors of correlation coefficients.
These two vectors are then correlated to one another to produce an 'autocorrelation.'
Low autocorrelation values are indicative of proteins whose interaction profiles are rewired between conditions, whereas high autocorrelation values indicate consistent elution profiles.

```{r}
# PXD014820
autocor1 = calculate_autocorrelation(chroms[['PXD014820|Ctrl']],
                                     chroms[['PXD014820|RNAseA']])
# PXD015406
autocor2 = calculate_autocorrelation(chroms[['PXD015406|Control']],
                                     chroms[['PXD015406|RNaseA']])
# inspect the output
str(autocor1)
```

By default, the `calculate_autocorrelation` function will compute the Pearson correlation, but the Spearman or Kendall correlation can be specified instead using the `cor_method` argument.
Other arguments to `calculate_autocorrelation` allow the user to remove proteins not detected in at least a certain number of fractions in either replicate (`min_fractions`), or to ignore correlations between protein pairs not detected in at least a certain number of overlapping fractions (`min_pairs`). 

Now, we combine the results from each experiment and print some of the most 'rewired' proteins.

```{r}
autocor_df = data.frame(accession = rep(c('PXD014820', 'PXD015406'), 
                                        c(length(autocor1), length(autocor2))),
                        protein = c(names(autocor1), names(autocor2)),
                        autocor = c(autocor1, autocor2) %>% unname())
autocor_df %>% arrange(autocor) %>% head()
```

Several of the top-ranking proteins, such as SRP14, SNRPA1, and SNRPN, are well-known RNA-binding proteins, providing an initial measure of confidence in the autocorrelation approach. 

To validate the method's ability to identify RNA-binding proteins more systematically, we now wish to perform a gene-set enrichment analysis of known RNA-binding proteins on the results of the autocorrelation analysis.
To achieve this, we begin by loading lists of known human RNA-binding proteins from the AtTRACT and RNABP databases.
(The raw datasets are preprocessed in the scripts `R/preprocessing/preprocess-AtTRACT.R` and `R/preprocessing/preprocess-RNABP.R`, respectively.)

```{r}
attract = readRDS("data/AtTRACT/AtTRACT.rds")
rbpdb = readRDS("data/RBPDB/RBPDB.rds")
rbps = bind_rows(mutate(attract, source = 'AtTRACT'),
                 mutate(rbpdb, source = 'RBPDB'))
sample_n(rbps, 5)
```

Here, the gene-set enrichment analysis is performed using functions from the package `fgsea` (Korotkevich et al., _bioRxiv_ 2019, doi: 10.1101/060012). 

```{r warning=FALSE, message=FALSE}
# convert RBPs to a list for fgsea
pathways = unstack(rbps[, c('Gene', 'source')])
str(pathways)
gsea = autocor_df %>%
  split(.$accession) %>% 
  # convert to vector for fgsea
  map(~ setNames(.x$autocor, .x$protein)) %>%
  # use only 10^5 permutations here
  map(~ fgsea(pathways, .x, nperm = 10^5, scoreType = 'neg')) %>%
  bind_rows(.id = 'accession')
gsea
```

Both RBP databases display a highly significant enrichment among 'rewired' proteins.
This can also be visualized by plotting the p-values: 

```{r fig.align='center', fig.width=1.5, fig.height=2.5}
pos = position_dodge(width = 0.6)
pal = c(colours.cafe322[5], darken(colours.cafe322[3], 1.1))
p = gsea %>%
  mutate(source = paste0(accession, '\nRNase vs. control')) %>%
  ggplot(aes(x = source, y = -log10(pval), color = pathway)) +
  geom_hline(aes(yintercept = -log10(0.05)), linetype = 'dotted', size = 0.3) +
  geom_errorbar(aes(ymin = 0, ymax = -log10(pval)), width = 0, position = pos) +
  geom_point(size = 0.9, position = pos) +
  scale_y_continuous(expression(-log[10](P)), limits = c(0, 6.5)) + 
  scale_color_manual(name = '', values = pal) +
  boxed_theme(size_sm = 5, size_lg = 6) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
p
```
